ggalluvial: Layered Grammar for Alluvial Plots

### 2020-03-24 paper/paper.rmd
simplify joss paper knit procedure
mv paper files to paper subdirectory

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "paper/figures/"
)
```

@@

Alluvial diagrams use stacked bar plots and variable-width ribbons to represent multi-dimensional or repeated-measures data comprising categorical or ordinal variables [@Rosvall2010;@Bojanowski2016]. The ggalluvial package extends the layered grammar of graphics of ggplot2 [@Wickham2016] to generate alluvial diagrams from tidy data [@Wickham2014].
The package makes two key contributions to the R ecosystem, which i summarize here and discuss in more detail below.

First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which imbues the resulting plots with greater meaning, most notably a cumulative axis, and imparts many combinatorial possibilities, e.g. with the text and point--line geoms of ggplot2. Meanwhile, the package gives users a high degree of control over underlying data transformations, and thereby over the messages conveyed in the plot, whereas other extensions are not so flexible.[^easy] The package also integrates alluvial plots to a tidyverse workflow [@Wickham2019] by recognizing two common tidy data formats and exporting data transformation functions that follow tidyr and dplyr conventions.

Second, ggalluvial cements some concepts and terms that are inconsistently applied by developers and users of Sankey, parallel sets, and alluvial diagrams.
For example, the diagrams that came to be known as parallel sets plots [@Kosara2006] were anticipated in the "start-up state" of an interactive generator of Sankey diagrams [@Riehmann2005], using wholly different terms for their graphical elements.
Since alluvial diagrams were introduced, implementations and applications have borrowed terminology from both previous diagram types. Instead, ggalluvial adopts a distinctive, geological nomenclature for its graphical elements. While this risks exacerbating confusion, the "alluvial plots" described here are, i think, distinctive enough from Sankey diagrams and parallel sets plots to warrant it. I hope that these conventions can serve as useful points of reference as these visualization tools converge toward common standards.

[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for expedited data exploration using alluvial plots.

@@

[The titular vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for more thorough illustrations. In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:

* _strata_, or stacked bar plots, located at each of multiple horizontal positions (_axes_ or _dimensions_)
* _alluvia_, ribbons through strata at different axes that encode individual cases or cohorts measured along each axis
* _lodes_, subdivisions of strata by their intersections with alluvia
* _flows_, segments of alluvia between adjacent axes (excluding lodes)

The layers of a ggplot2 graphic are formed by pairing stats (statistical transformations of the input data) with geoms (geometric mappings from the transformed data); while every stat and geom has a conventional default, alternative pairings are the primary source of combinatorial richness for this layered grammar.
The following alluvial plot depicts several meaningful stat--geom combinations that appear in the documentation. Default pairings, other within-package ("alluvial") pairings, and pairings of alluvial stats with other geoms are differentiated by `fill` color:

```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}
library(ggalluvial)
combos <- data.frame(
  stat = c(
    rep("stratum", 2),
    rep("alluvium", 4),
    "flow",
    # text geom
    c("stratum", "alluvium", "flow"),
    # errorbar and pointrange geoms
    "stratum", "alluvium", "flow"
  ),
  geom = c(
    c("stratum", "lode"),
    c("stratum", "alluvium", "lode", "flow"),
    "flow",
    rep("text", 3),
    "errorbar", "pointrange", "pointrange"
  ),
  pairing = c(
    c("default", "alluvial"),
    c("alluvial", "default", "default", "alluvial"),
    "default",
    rep("external", 6)
  )
)
levs <- c("stratum", "alluvium", "lode", "flow",
          "text", "errorbar", "pointrange")
combos <- transform(
  combos,
  stat = factor(stat, levs),
  geom = factor(geom, levs),
  pairing = factor(pairing, c("default", "alluvial", "external"))
)
ggplot(combos, aes(axis1 = stat, axis2 = geom)) +
  scale_x_discrete(limits = c("stat", "geom"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = pairing)) +
  scale_fill_brewer(type = "qual") +
  geom_stratum() +
  geom_text(stat = "stratum", infer.label = TRUE) +
  theme_minimal()
```

The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format. Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions. The long (one row per lode) and wide (one row per alluvium) formats are also detailed in the vignette, and are related to each other by the pivot operations of tidyr. Because the alluvial geoms are highly specialized to these stats, no pairings with outside stats are currently supported.

Most of the stat parameters control how the strata at each axis, and the lodes within each stratum, are ordered vertically. By default, these orderings are independent of differentiation aesthetics, so that layers are consistent within and across plots unless otherwise specified. [An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of each of these parameters. For convenience, they can also be set as global options.

While questions and requests by users have prompted improvements throughout ggalluvial's development, the current version is expected to be feature complete, with one exception: alternative curves will be implemented for the `alluvium` and `flow` geoms, controlled by a new parameter. The current splines will remain the default, so that the output of existing code will not change. This is expected to be the last change before version `1.0`.

@@

Statistical graphics of multivariate categorical data tend to take one of two forms: ordinations of continuous-valued transformations (e.g. residual plots from multiple regression; ternary plots from tripartition data; biplots from correspondence analysis and factor analysis) and arrangements of length- or area-encoded shapes (e.g. mosaic plots [@Friendly2002]; product plots [@Wickham2011]; stacked area charts).
In particular, parallel sets plots, adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990], represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].

Meanwhile, visualizations of flow processes have long encoded magnitudes, such as volumes and counts, as ribbon widths, and this broad genre have come to be called Sankey diagrams [@Schmidt2008]; see <http://www.sankey-diagrams.com/> for many examples.
A widely-used and -implemented subtype of Sankey diagram encodes longitudinal categorical data as rectangular nodes, representing categories, threaded by ribbons that encode the trajectories and magnitudes of cases [@Riehmann2005].

Most recently, @Rosvall2010 proposed "alluvial diagrams" to visualize changes in cluster membership among cases over successive cross-sections, with cohorts of equivalent cases encoded as ribbons intersecting stacked bar plots at each time point. These diagrams have since been used to represent a wide range of discrete-valued data, over repeated measures or multiple dimensions.

Several R packages, including ggplot2 extensions, have been developed to generate these types of data visualizations. Here are the most relevant for this discussion:

- alluvial for alluvial diagrams in base R [@Bojanowski2016]
- ggforce for parallel sets plots in ggplot2 [@Pedersen2019]
- ggparallel and ggpcp for parallel sets, hammock, and common angle plots in ggplot2 [@Hofmann2013;@Ge2019]
- ggalluvial for alluvial diagrams in ggplot2 [@Brunson2019]

While Sankey diagrams, parallel sets plots, and alluvial diagrams have accrued their own characteristic use cases, the terms are often used interchangeably, and there is currently no consensus on what features, if any, are distinctive to each type. Moreover, these visualization techniques in practice vary widely in their flexibility or rigidity---some rely on non-deterministic layout algorithms and can even be interactively manipulated, while others are wholly determined by the data; most give the user limited control through parameter settings.

The diagrams produced by ggalluvial are statistical graphics in the sense that they communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006]. The package adopts for them the less common term _alluvial plot_.[^plots]
These plots are distinct from Sankey diagrams (they exhibit a much narrower range of behavior, e.g. no cycles, and do not necessarily represent flow) and constitute a subtype of parallel sets plots distinguished by two features: a strict order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a meaningful positional dimension along which these elements are stacked, which precludes inserting gaps between them (as is typical of most diagrams of each type), features shared only by ggparallel. The plot elements (strata, either alluvia or flows, and optionally lodes or others as discussed above) are rendered a separate layers, following the additive (`+`) syntax of ggplot2, a feature shared only by ggpcp.

[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.

@@

ggalluvial has seen widespread use in research and scholarship.
While many of these uses would be served equally well by other parallel sets plots or Sankey diagrams, i showcase here three settings to which alluvial plots seem exceptionally well-suited: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.
To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.

**Repeated ordinal measures data:**
Parallel sets plots and Sankey diagrams are commonly used to represent repeated categorical measures data. Most implementations stack each bar plot in order of name or of size, though some follow hierarchies provided by the user, often with vertical space between category for easy visual discrimination. In contrast, ordinal variables are more appropriately stacked in their own intrinsic and consistent order, and, when the number of categories changes from time to time, vertical separations can obscure whether the cases and cohorts changed in number as well. A use case by @Schlotter2019, to represent changes (or not) in patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in order across axes (in this case preference ranks rather than time points).

**Incomplete longitudinal data:**
Alluvial plots very clearly delineate times at which longitudinal data are censored, discontinued, or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected (CDI) patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this dominant type. While all 32 patients had at least two samples taken, only 3 had four, so that the bar plots shortened with each time point.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation rather than omitting them entirely from the plot.
<!--
Even in cases of intermediate missingness, when a case is lost, or its value diminished to zero, between two time points at which it is positively measured, alluvial diagrams can preserve the continuity of such cases by shrinking their ribbon widths to zero at the intermediate point. (See the example in the `stat_alluvium()` examples using the babynames package.)
-->

**Signed categorical data:**
@Edwards2019 recently produced a novel alluvial plot to represent changes in ownership category (owner-operated, corporate, etc.) of owners in a halibut fishery. Not only the categorical distribution but the total number of owners changed from year to year as exeters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exeters at each year, the authors affixed a negative stratum for the exeter category to each bar plot, below the horizontal axis. Such a feature has no analogue in Sankey diagrams or parallel sets plots (which do not commonly employ a cumulative axis) but potentially wide-ranging applications: Bar plots may include both "positive" and "negative" bars to represent definitionally negative groups, such as revenues versus deficits, or to contrast the bars divided along a binary variable such as gender across age groups in a population (so-called "pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such bar plots, even when cases may change sign. Future use cases may demonstrate the practical utility of this functionality.

@@

I am grateful to many users for their feedback on every version of this package. Development benefitted from the use of resources and support of colleagues at UConn Health, and i have been supported in part by T90 training grant from the National Institute of Dental and Craniofacial Research (5T90DE021989-07).

>>>>

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "paper/"
)
```

@@

Alluvial diagrams use stacked bar plots and variable-width ribbons to represent multi-dimensional or repeated-measures data comprising categorical or ordinal variables [@Rosvall2010;@Bojanowski2016]. The ggalluvial package extends the layered grammar of graphics of ggplot2 [@Wickham2016] to generate alluvial diagrams from tidy data [@Wickham2014].

The package makes two key contributions to the R ecosystem.
First, ggalluvial anchors the imprecise notion of an alluvial diagram to the rigid grammar of graphics [@Wilkinson2006], which lends the plots more precise meaning and opens up many combinatorial possibilities.
Second, ggalluvial adopts a distinctive geological nomenclature to distinguish "alluvial plots" and their graphical elements from Sankey diagrams and parallel sets plots, which I hope prove useful as these visualization tools converge toward common standards.

@@

[The primary vignette](http://corybrunson.github.io/ggalluvial/articles/ggalluvial.html) thoroughly describes and illustrates the functionality of ggalluvial, and the reader is encouraged to browse [the package documentation](http://corybrunson.github.io/ggalluvial/reference/index.html) for comprehensive examples. In brief, the package contains stat and geom functions to add the following layers to a ggplot2 object:

* _strata_, or stacked bar plots, located in parallel along a (plotting) axis of (variable) _axes_ or _dimensions_
* _alluvia_, ribbons through strata that connect the categories of individual cases or cohorts at different axes
* _lodes_, subdivisions of strata by their intersections with alluvia
* _flows_, segments of alluvia between strata

Figure \ref{fig:example} illustrates these and other plot elements by visualizing changes in several students' curricula (based on their declared majors) across several academic terms.
Each axis corresponds to an odd-valued term (1 through 15), at which the students are grouped into strata according to their curricula---Art History, Ceramic, etc. The individual students can be tracked from term to term along their alluvia: for instance, one student started out in Digital Art, encoded by the blue ribbon, but had switched to Painting by the 11th term, where the ribbon turns pink. The partially transparent flows are colored according to their originating (not their terminating) terms, and the lodes where they intersect the strata are obscured by the solid-colored strata themselves. When a student's curriculum is unknown, they are grouped into the "missing" (`NA`) stratum, which is weighted negatively in this example.

```{r figure, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8, fig.cap="\\label{fig:example}Alluvial plot of changes in curricula by a cohort of art students"}
library(ggalluvial)
data(majors)
majors$curriculum <- as.factor(majors$curriculum)
levels(majors$semester) <- gsub("CURR", "Term ", levels(majors$semester))
arr <- arrow(angle = 15, type = "closed", length = unit(.03, "npc"))
ggplot(majors,
       aes(x = semester, stratum = curriculum, alluvium = student,
           fill = curriculum)) +
  scale_fill_brewer(type = "qual", palette = "Set2", name = "Curriculum") +
  scale_y_continuous(breaks = seq(-2, 10, 2)) +
  geom_flow(stat = "alluvium", lode.guidance = "frontback",
            color = "darkgray", negate.strata = NA) +
  geom_stratum(negate.strata = NA) +
  theme(legend.position = "right") +
  labs(x = NULL, y = "Number of students") +
  # annotate axes
  annotate("segment", x = 2.33, y = -1.33, xend = 2, yend = -.33, arrow = arr) +
  annotate("segment", x = 3, y = -1.33, xend = 3, yend = -.33, arrow = arr) +
  annotate("segment", x = 3.67, y = -1.33, xend = 4, yend = -.33, arrow = arr) +
  annotate("text", x = 3, y = -1.67, label = "axes / dimensions", size = 4) +
  # annotate strata
  annotate("segment", x = 1.5, xend = 1, y = 3.5, yend = 1.5, arrow = arr) +
  annotate("segment", x = 1.5, xend = 1, y = 4, yend = 3.5, arrow = arr) +
  annotate("segment", x = 1.5, xend = 1, y = 4.5, yend = 6, arrow = arr) +
  annotate("text", x = 1.67, y = 4, label = "strata", angle = 90, size = 4) +
  # annotate alluvia
  annotate("segment", x = 4.33, y = 6.5, xend = 1, yend = 8.5, arrow = arr) +
  annotate("segment", x = 5, y = 6.5, xend = 4.5, yend = 9, arrow = arr) +
  annotate("segment", x = 5.67, y = 6.5, xend = 7.5, yend = 7, arrow = arr) +
  annotate("text", x = 5, y = 6.17, label = "one alluvium", size = 4) +
  # annotate flows
  annotate("segment", x = 5.33, y = 2.5, xend = 4.25, yend = 0.5, arrow = arr) +
  annotate("segment", x = 5.33, y = 3, xend = 4.25, yend = 2.5, arrow = arr) +
  annotate("segment", x = 5.33, y = 3.5, xend = 3.75, yend = 4.5, arrow = arr) +
  annotate("text", x = 5.5, y = 3, label = "flows", angle = 90, size = 4) +
  # annotate lodes
  annotate("segment", x = 7.5, y = 3, xend = 8, yend = 1.5, arrow = arr) +
  annotate("segment", x = 7.5, y = 3.5, xend = 8, yend = 3.5, arrow = arr) +
  annotate("segment", x = 7.5, y = 4, xend = 8, yend = 5.5, arrow = arr) +
  annotate("text", x = 7.33, y = 3.5, label = "lodes", angle = -90, size = 4)
```

Plot layers are formed by pairing stats (statistical transformations) with geoms (mappings to graphical elements and properties); while every stat and geom has a conventional default, alternative grammatical pairings provide combinatorial richness to plotting possibilities. In the above example, the alluvium geom was paired with the flow stat, so that the flows of each alluvium could change color across the axes.
Other meaningful stat--geom combinations can be found in the documentation, including pairings of the three alluvial stats (stratum, alluvium, and flow) with the text, errorbar, and pointrange geoms.

Alluvial layers can interpret tidy data in either of two formats: long (one row per lode) and wide (one row per alluvium). These are related by the pivot operations of tidyr [@Wickham2019] and can be toggled between using the custom functions `to_lodes_form()` and `to_alluvia_form()`.
The alluvial stats require custom aesthetics---either `stratum` and/or `alluvium` in combination with `x`, if the data are in long format, or some number of axis specifications (`axis1`, `axis2`, etc.), if the data are in wide format.[^warnings] Because the alluvial geoms are specialized to these stats, no pairings with outside stats are currently supported.

[^warnings]: Because these aesthetics are not recognized by ggplot2, they produce warnings under some conditions.

Most of the stat parameters control how the strata at each axis, and the lodes within each stratum, are ordered vertically. By default, these orderings are independent of differentiation aesthetics, so that layers are consistent within and across plots unless otherwise specified. [An auxiliary vignette](http://corybrunson.github.io/ggalluvial/articles/order-rectangles.html) details the effects of these parameters. They can also be set as global options.

@@

Visualizations of flow processes have long encoded magnitudes as ribbon widths, constituting a type called Sankey diagrams [@Schmidt2008].
A widely-used subtype for longitudinal categorical data represent categories as nodes threaded by edges that represent the trajectories and magnitudes of cases [@Riehmann2005].
Their design anticipated parallel sets plots, which were adapted from parallel coordinates plots [@Inselberg1987;@Wegman1990] to visualize multivariate categorical data, and which represent cohorts of equivalent cases as ribbons connecting categories represented as boxes [@Kosara2006].
These in turn anticipated "alluvial diagrams", proposed to visualize changes in case memberships across successive cross-sections [@Rosvall2010].
Several R packages have been developed to generate diagrams of these types, including riverplot [@Weiner2017], networkD3 [@Allaire2017], sankey [@Csardi2017], alluvial [@Bojanowski2016], ggparallel [@Hofmann2013], ggforce [@Pedersen2019], ggalluvial [@Brunson2019], and ggpcp [@Ge2019].

Sankey, parallel sets, and alluvial diagrams are often conflated, and there is currently no consensus on what features are distinctive to each type.
Moreover, their graphical elements go by a variety of names, often interchangeably.
In order to more clearly describe the features of ggalluvial in relation to similar packages, I have found it useful to adopt a careful demarcation among these diagram types.

_Statistical graphics_ (here also simply called "plots") are diagrams that communicate statistical information using graphical methods [@Friendly2005] and, more narrowly, are uniquely determined from data by a fixed set of plotting rules [@Wilkinson2006].
By design, graphics produced by ggplot2 extensions are plots: The stat, geom, and other layers of a ggplot object exactly reproduce a graphic from data (under the same parameter settings).[^exceptions]
Sankey diagrams are much more flexible.
The earliest engine efficiency diagrams in this tradition could take a variety of forms to depict the same energy flow and were differently annotated for different audiences [@Schmidt2008].
Software implementations may use heuristic algorithms to position their graphical elements [@Allaire2017;@Csardi2017] or enable users to manually, even interactively, adjust them [@Riehmann2005;@Allaire2017;@Weiner2017].
Paradoxically, Sankey diagrams are overwhelmingly used to represent flow, whereas the aforecited ggplot2 extensions are used to visualize a wide variety of data types.
Arguably, these extensions are better understood as producing a different type of diagram.

[^exceptions]: There are exceptions. For example, the jitter geom in ggplot2 introduces randomness to symbol positions, and the repel geoms of ggrepel [@Slowikowski2019] use heuristic algorithms to position text.

Parallel sets plots might be viewed as a subtype of Sankey diagram with the following features: Ribbons proceed monotonically along one dimension, and every ribbon encounters a box at every axis. These graphical constraints correspond to combinatorial constraints on the data, which amount to an id--key--value structure in which every id--key pair takes exactly one value (possibly zero or missing, and optionally weighted).
In this sense, the plots produced by the ggplot2 extensions (and by the alluvial package) are parallel sets plots: Cohorts are partitioned into categories at each axis and connected by ribbons whose widths encode their magnitudes.[^hammock]

The plots produced still vary---in the shapes of ribbons, the arrangements of boxes, and the presence of gaps between boxes at the same axis.
The exceptional geoms of ggparallel each offer common-angle as well as linear ribbons. Those of alluvial, ggforce, ggalluvial, and ggpcp offer one-parameter families that interpolate between straight and x-spline ribbons.[^curves]
The stats vertically arrange the elements (boxes and ribbons) at each axis.
These distinct elements are rendered by separate layers in ggforce, ggalluvial, and ggpcp, following the additive (`+`) syntax of ggplot2.
ggalluvial provides more levers of control over the statistical transformations, thereby over the messages conveyed by the plot, than the other packages.[^easy]

[^curves]: Several alternative curves, based on @Shaffer2019, are in development.
[^easy]: Indeed, the dependency package easyalluvial [@Koneswarakantha2019] was built on top of ggalluvial to exchange much of this flexibility for more expedient data exploration.

The ggalluvial package adopts the term _alluvial plot_ for the subtype of parallel sets plots it produces, with the geological terminology introduced above.[^plots]
These alluvial plots are distinguished by two features: a prescribed order on the stacked elements at each axis, including both the values of the discrete variables and the ribbons connecting cases or cohorts between them; and a real-valued plotting dimension perpendicular to that of flow, along which these elements are stacked, so that gaps between them are precluded.
In combination, these features confer greater meaning on the second plotting dimension.

The first feature is shared by the other packages but is not essential to parallel sets plots; such plots could, for example, arrange boxes corresponding to repeated categorical decompositions differently at different axes. While most of the packages separate boxes at each axis with gaps, these can be reduced to zero, so that each package can create alluvial plots. (ggparallel and ggalluvial alone _only_ produce alluvial plots.)
These features are particularly important to some applications and, in my view, can fundamentally change the way a plot is interpreted.
It is for this reason that I believe the new typology and terminology are warranted.

[^hammock]: The possible exceptions are the hammock plots and common angle plots of ggparallel, which are contrasted with a stricter definition of parallel sets plots than I use here, in which ribbons are straight, their widths aggregate to box widths, and they meet without overlap at the sides of boxes, partitioning them [@Hofmann2013].
[^plots]: This has the unfortunate side effect of conflating search results from the geology literature.

@@

While most uses might be served equally well by other parallel sets plots or Sankey diagrams, alluvial plots seem exceptionally well-suited to three settings: repeated ordinal measures data, incomplete longitudinal data, and signed categorical data.[^opinion]

[^opinion]: To be sure, this is a subjective assessment that may be refuted by visualization effectiveness research.

**Repeated ordinal measures data.**
Most Sankey, parallel sets, and alluvial implementations stack each bar plot in order of name or of size (though some follow user-provided hierarchies), and most insert gaps between categories for easy visual discrimination. Ordinal variables are most appropriately stacked in their own intrinsic and consistent order and, when the number of categories (hence of gaps) changes from axis to axis, vertical separations can obscure whether magnitude totals changed as well. A use case by @Schlotter2019, to represent patients' physical limitations following an investigational right heart valve repair technique, illustrates the use of an ordinal stratum variable (a heart failure functional classification). Another, by @North2019, to represent ranked preferences among several definitions of veganism by survey respondents, illustrates the importance of consistency in their order. In both cases, the fixed heights of the bar plots conveyed that no individuals were lost to follow-up.

**Incomplete longitudinal data.**
Alluvial plots clearly indicate times at which longitudinal data are censored or otherwise missing: Certain strata, or the alluvia or flows connecting them, are present at one time point but absent at a previous or future one.
@Seekatz2018 use this feature to include in one alluvial plot a sample of _Clostridium difficile_--infected patients who had their infections ribotyped at multiple times. Patients were classified by dominant ribotype, and the alluvial plot showcased variability in this classification. While all 32 patients had at least two samples taken, only 3 had four, communicated by the shortening of the bar plots along the main dimension.
@Sjoding2019 use a similar plot to trace patient groups receiving mechanical ventilation based on discretized tidal volumes, including a grey stratum for patients discontinued from intubation.

**Signed categorical data.**
@Edwards2019 produced a novel alluvial plot to represent changes in ownership category of owners in a halibut fishery. The total number of owners changed from year to year as exiters were not exactly matched by new entrants. In order to depict an accurate total but include both new entrants and exiters at each year, the authors affixed a negative stratum for the exiter category to each bar plot.[^negative] Such a feature has no analogue in Sankey diagrams or parallel sets plots but potentially wide-ranging applications: Bar plots may use "positive" and "negative" bars to represent signed categories, such as contributors to revenue versus deficit, or to contrast the bars divided along a binary variable such as gender across age groups in a population ("pyramid plots"). Alluvial plots provide a way to track cases and cohorts across such graphics, even when cases change sign. Future applications may demonstrate additional uses for this functionality.

[^negative]: The authors should be credited with this innovation, which I only implemented in ggalluvial after learning about their workaround to create it using a previous version.

@@

I am grateful to many users for their feedback on every version of this package. Development benefitted from the use of resources and support of colleagues at UConn Health, and I have been supported in part by T90 training grant 5T90DE021989-07 from the National Institute of Dental and Craniofacial Research.
